---
title: "Simulating Discrete Choice Experiments with simulateDCE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Discrete Choice Experiments with simulateDCE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Build instructions
1. Create vignette using usethis::use_vignette("my-vignette")
2. After making changes, run devtools::build_vignettes()
3. Rebuild using devtools::install(build_vignettes = TRUE)
4. Check with browseVignettes()

If the vignette does not appear on GitHub, rendered files may have landed in /doc
instead of /inst/doc. To fix:
  tools::buildVignettes(dir = ".", tangle = TRUE)
  dir.create("inst/doc")
  file.copy(dir("vignettes", full.names = TRUE), "inst/doc", overwrite = TRUE)
-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `simulateDCE` package provides a workflow for simulating discrete choice
experiment (DCE) datasets. The intended use is to evaluate the statistical
power of a proposed experimental design *before* collecting real data: by
simulating many artificial datasets from a design and estimating a model on
each one, you can judge whether your design and sample size are likely to
recover the parameters you care about.

This vignette walks through the full workflow using a community-supported
agriculture (CSA) example design that ships with the package.

```{r setup}
library(simulateDCE)
library(rlang)
library(formula.tools)
```

---

# Step 1 — Specify the utility functions and beta coefficients

A DCE asks respondents to choose between alternatives that differ on
a set of attributes. The analyst postulates a **utility function** for
each alternative — a linear combination of attributes weighted by beta
coefficients. The simulation uses these functions to calculate the
deterministic part of utility; a Gumbel-distributed error term is then
added to generate actual choices.

Utility functions are written as R formulas. The left-hand side names the
utility term (`V.1`, `V.2`, …); the right-hand side is an expression in
the design attributes and beta coefficients. All beta names must start
with `b` by convention, and they must match the names in the `bcoeff` list
exactly.

```{r initialize}
# Beta coefficients: hypothesised true parameter values
bcoeff <- list(
  bx1 = -0.1,
  bx2 = -0.1,
  bx3 = -0.05,
  bx4 = -0.025
)

# Utility functions — one sub-list per decision group (here just one group)
ul <- list(u1 = list(
  v1 = V.1 ~ bx1 * alt1.x1 + bx2 * alt1.x2 + bx3 * alt1.x3 + bx4 * alt1.x4,
  v2 = V.2 ~ bx1 * alt2.x1 + bx2 * alt2.x2 + bx3 * alt2.x3 + bx4 * alt2.x4,
  v3 = V.3 ~ -5   # opt-out alternative with fixed utility
))
```

The `bcoeff` list maps each coefficient name to a **fixed scalar** — the same
value is used for every simulated respondent. This corresponds to a standard
multinomial logit (MNL) model. See the section on random parameters below for
the mixed logit extension.

---

# Step 2 — Point to the experimental design

`sim_all()` reads one or more design files from a folder and runs the
simulation for each design it finds. You need to tell it:

- **`designpath`**: the folder containing the design file(s).
- **`designtype`**: the software used to create the design (`"spdesign"` or
  `"ngene"`). If you omit this, the package will try to detect it
  automatically.

```{r other}
designpath <- system.file("extdata", "CSA", "linear", package = "simulateDCE")
# For your own designs, use the folder path directly, e.g.:
# designpath <- "Projects/CSA/Designs/"

notes    <- "No Heuristics"
resps    <- 240   # number of respondents per simulated dataset
nosim    <- 2     # number of simulation runs (use >= 500 for real analyses)
destype  <- "spdesign"
```

The `resps` argument controls how large each simulated sample is. The
`nosim` argument controls how many independent datasets are generated and
estimated; the summary statistics across those replications tell you about
the expected performance of your design. For exploratory use a small number
is fine; for publication-quality power analyses use at least 500.

---

# Step 3 — Set a random seed (optional but recommended)

Several steps in the simulation draw random numbers — Gumbel errors, and
any random parameters you specify. Setting a seed makes results
reproducible.

```{r random}
set.seed(3393)
```

---

# Step 4 — Run the simulation

`sim_all()` ties everything together. It reads every design file in
`designpath`, expands it across `resps` respondents, simulates choices
`nosim` times, and (by default) estimates a logit model on each replicated
dataset.

```{r output}
csa <- simulateDCE::sim_all(
  nosim      = nosim,
  resps      = resps,
  designtype = destype,
  designpath = designpath,
  u          = ul,
  bcoeff     = bcoeff
)
```

---

# Step 5 — Inspect the results

`sim_all()` returns a nested list. The top-level names correspond to each
design file found in `designpath`. Inside each design, the key elements are:

| Element | Contents |
|---------|----------|
| `coefs` | Data frame of estimated coefficients from every simulation run |
| `summary` | Summary statistics (mean, SD, % significant) across runs |

```{r accessOutput}
# Names of the design-level results
topLevelResults <- names(csa[sapply(csa, is.list)])
print(topLevelResults)

# Coefficients and summary for the first design
simulationCoeff <- csa[[1]]$coefs
coeffSummary    <- csa[[1]]$summary

print(simulationCoeff)
print(coeffSummary)
```

The `summary` table is the main diagnostic. Look at the `mean` column to
check that the estimated coefficients are close to the true values you
specified in `bcoeff`, and at the `% significant` column to assess power.

You can persist the results with:
```r
saveRDS(csa, file = "results/csa_simulation.RDS")
```

---

# Extension — Random parameters (mixed logit)

The fixed-coefficient MNL model above assumes all respondents share the
same preferences. In reality, preferences vary. A **mixed logit** model
accounts for this by drawing each respondent's coefficients from a
distribution rather than using a single value.

In `simulateDCE`, you switch to random parameters simply by replacing a
scalar in `bcoeff` with a named list that specifies the distribution.
Supported distributions are:

| `dist` value | Required parameters | Notes |
|---|---|---|
| `"normal"` | `mean`, `sd` | Unbounded; suitable for most attributes |
| `"lognormal"` | `meanlog`, `sdlog` | Always positive |
| `"neg_lognormal"` | `meanlog`, `sdlog` | Always negative; useful for cost |
| `"uniform"` | `min`, `max` | Bounded |
| `"triangular"` | `min`, `max`, `mode` | Bounded with a peak |

Fixed and random parameters can be mixed freely within `bcoeff`.

```{r random-params}
bcoeff_mixed <- list(
  bx1 = list(dist = "normal", mean = -0.1, sd = 0.05),  # random: normally distributed
  bx2 = list(dist = "normal", mean = -0.1, sd = 0.05),  # random: normally distributed
  bx3 = -0.05,                                           # fixed: same for all respondents
  bx4 = -0.025                                           # fixed: same for all respondents
)
```

Pass this mixed `bcoeff` to `sim_all()` exactly as before. The function
will automatically detect the random parameters, draw respondent-level
values from the specified distributions, and merge them into the dataset
before computing utilities. The parameter summary printed at startup
confirms how each coefficient has been interpreted:

```{r run-mixed, eval = FALSE}
set.seed(3393)

csa_mixed <- simulateDCE::sim_all(
  nosim      = nosim,
  resps      = resps,
  designtype = destype,
  designpath = designpath,
  u          = ul,
  bcoeff     = bcoeff_mixed
)
```

> **Note:** When using random parameters, the output dataset will contain an
> extra column for each random coefficient (one draw per respondent). These
> columns are visible in the raw simulated data but are not part of the
> estimated model output.

You can also call `make_rand_params()` directly to inspect the draws
before running a full simulation:

```{r make-rand}
draws <- make_rand_params(bcoeff_mixed, n_resp = 10)
print(draws)
```

Each row is one respondent. Fixed parameters repeat the same value;
random parameters vary across rows according to the chosen distribution.
