---
title: "Decision groups and heuristics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Decision groups and heuristics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(simulateDCE)
set.seed(2210)
```

Random utility maximisation says respondents weigh up every attribute and pick
the best alternative. Plenty of them do something else. Some ignore an attribute
entirely. Some look only at price. Some always take the status quo. Decision
groups let you simulate a sample where different people follow different rules,
then see what your model makes of it.

The question this answers is uncomfortable and worth asking: if a third of my
respondents ignore the attribute my paper is about, will I notice?

## How it works

Two arguments work together.

`u` is a list of lists. Each inner list is one decision rule, written as a full
set of utility functions. `decisiongroups` is a vector of cumulative shares
saying how the sample splits between them. It starts at 0, ends at 1, and needs
one more element than there are groups.

```
decisiongroups = c(0, 1)          one group, everybody
decisiongroups = c(0, 0.7, 1)     70% follow u1, 30% follow u2
decisiongroups = c(0, 0.5, 0.8, 1) 50% u1, 30% u2, 20% u3
```

Whole respondents are assigned to a group. A respondent never answers some of
their choice tasks under one rule and the rest under another.

## A sample with attribute non-attendance

Three attributes this time, so I build the design from the full factorial and pair
each profile with a randomly chosen other one:

```{r design}
set.seed(4711)

profiles <- expand.grid(price = c(2, 4, 6, 8), quality = c(0, 1), origin = c(0, 1))
n <- nrow(profiles)
partner <- sample(n)

design <- data.frame(
  Choice.situation = seq_len(n),
  Block            = rep(1:2, each = n / 2),
  alt1.price       = profiles$price,
  alt1.quality     = profiles$quality,
  alt1.origin      = profiles$origin,
  alt2.price       = profiles$price[partner],
  alt2.quality     = profiles$quality[partner],
  alt2.origin      = profiles$origin[partner]
)

design_dir <- file.path(tempdir(), "dg_designs")
dir.create(design_dir, showWarnings = FALSE)
saveRDS(design, file.path(design_dir, "design.rds"))

bcoeff <- list(bprice = -0.3, bquality = 0.5, borigin = 0.4)
```

With two alternatives, what a logit can identify comes from the *differences*
between them, so it is worth checking those before simulating anything:

```{r check-design}
diffs <- with(design, data.frame(
  price   = alt1.price - alt2.price,
  quality = alt1.quality - alt2.quality,
  origin  = alt1.origin - alt2.origin
))

c(
  rank              = qr(as.matrix(diffs))$rank,   # must equal the number of attributes
  distinct_patterns = nrow(unique(diffs)),
  largest_cor       = max(abs(cor(diffs)[upper.tri(cor(diffs))]))
)
```

Rank 3 for 3 attributes means all three are identified. A dozen distinct
difference patterns and modest correlations mean they are identified comfortably
rather than only just. A design whose attribute differences are collinear, or that
offers only a handful of patterns, produces nonsense estimates and the occasional
run that walks off to infinity while still reporting successful convergence. That
is worth 30 seconds of checking, whether the design came from Ngene or from your
own head.

The first rule is ordinary RUM over all three attributes. The second ignores
origin: the term is simply absent from the utility functions.

```{r groups}
ul <- list(
  # attends to everything
  rum = list(
    v1 = V.1 ~ bprice * alt1.price + bquality * alt1.quality + borigin * alt1.origin,
    v2 = V.2 ~ bprice * alt2.price + bquality * alt2.quality + borigin * alt2.origin
  ),
  # ignores origin
  ignores_origin = list(
    v1 = V.1 ~ bprice * alt1.price + bquality * alt1.quality,
    v2 = V.2 ~ bprice * alt2.price + bquality * alt2.quality
  )
)
```

Note that both groups use the same `bcoeff`. A group differs by which terms
appear in its utility function, not by having its own coefficient values. If you
want a group with a different price sensitivity, give it a different coefficient
name and add that name to `bcoeff`.

## What non-attendance does to the estimates

```{r run}
share_ignoring <- c(0, 0.25, 0.5, 0.75)

result <- vapply(share_ignoring, function(share) {
  set.seed(88)
  dg <- if (share == 0) c(0, 1) else c(0, 1 - share, 1)
  u  <- if (share == 0) ul["rum"] else ul

  res <- sim_all(
    nosim = 40, resps = 300, designpath = design_dir,
    u = u, bcoeff = bcoeff, decisiongroups = dg,
    mode = "sequential", verbose = 0
  )
  sa <- res$summaryall
  est <- function(p) sa$design.mean[sa$parname == p]
  c(bprice = est("bprice"), bquality = est("bquality"), borigin = est("borigin"))
}, numeric(3))

colnames(result) <- paste0(100 * share_ignoring, "% ignore origin")
round(result, 3)
```

The origin coefficient collapses almost exactly in proportion to the share of
respondents who attend to it: three quarters of $0.4$, then half, then a quarter.
The estimate is an average over people who care and people who do not. The other
two coefficients are barely touched, because non-attendance is specific to one
attribute.

That is the trap. A halved `borigin` looks exactly like a real preference that is
weaker than you thought. Nothing in the output says "half your sample was not
reading this attribute". Simulating the heuristic first is what tells you what
size of estimate is consistent with what share of non-attendance.

## Power under a heuristic

The same thing shows up as lost power:

```{r power}
power <- vapply(share_ignoring, function(share) {
  set.seed(88)
  dg <- if (share == 0) c(0, 1) else c(0, 1 - share, 1)
  u  <- if (share == 0) ul["rum"] else ul

  res <- sim_all(
    nosim = 40, resps = 300, designpath = design_dir,
    u = u, bcoeff = bcoeff, decisiongroups = dg,
    mode = "sequential", verbose = 0
  )
  res$powa_by_par$design
}, numeric(3))

colnames(power) <- paste0(100 * share_ignoring, "%")
power
```

If you are powering a study on the origin attribute and you suspect
non-attendance, this table is the honest basis for the sample size you ask for.

## Other heuristics

The mechanism is general. Anything you can write as a utility function is a
decision rule.

**Price only.** Drop every other term:

```{r lexicographic}
ul_price_only <- list(
  rum = ul$rum,
  price_only = list(
    v1 = V.1 ~ bprice * alt1.price,
    v2 = V.2 ~ bprice * alt2.price
  )
)
```

**A stronger weight on one attribute.** Multiply the term, or introduce a
separate coefficient:

```{r stronger}
bcoeff_two <- c(bcoeff, list(bprice_sensitive = -0.9))

ul_sensitive <- list(
  normal = ul$rum,
  sensitive = list(
    v1 = V.1 ~ bprice_sensitive * alt1.price + bquality * alt1.quality + borigin * alt1.origin,
    v2 = V.2 ~ bprice_sensitive * alt2.price + bquality * alt2.quality + borigin * alt2.origin
  )
)
```

**Random choice.** Give every alternative the same utility, so only the Gumbel
error decides:

```{r random}
ul_random <- list(
  rum = ul$rum,
  coin_flip = list(v1 = V.1 ~ 0, v2 = V.2 ~ 0)
)

set.seed(3)
res <- sim_all(
  nosim = 20, resps = 300, designpath = design_dir,
  u = ul_random, bcoeff = bcoeff, decisiongroups = c(0, 0.8, 1),
  mode = "sequential", verbose = 0
)
sa <- res$summaryall
round(sa$design.mean[!grepl("^rob_pval0_", sa$parname)], 3)
```

With 20% of the sample choosing at random, all three coefficients shrink by
roughly the same proportion. That is the signature of noise, and it is different
from non-attendance, which pulls one coefficient down and leaves the others where
they were. If a real dataset gives you one suspiciously small coefficient, that
points at non-attendance; if everything is uniformly small, that points at noise.
Simulating both patterns first is how you learn to tell them apart.

## Checking the split

Set `estimate = FALSE` to see the group assignment itself:

```{r check}
sim_only <- sim_all(
  nosim = 1, resps = 300, designpath = design_dir,
  u = ul, bcoeff = bcoeff, decisiongroups = c(0, 0.7, 1),
  estimate = FALSE, mode = "sequential", verbose = 0
)

d <- sim_only$design[[1]]

# respondents per group, not rows per group
table(d$group[match(unique(d$ID), d$ID)])

# nobody straddles two groups
all(tapply(d$group, d$ID, function(x) length(unique(x))) == 1)
```

The `group` column is part of the simulated data, so it also gives you a
benchmark: you know the true class membership, which is exactly what a latent
class model has to recover. Simulating with known groups and then fitting a
latent class model to the result is a good way to find out how well that model
can do on your design.
