---
title: "Reading designs from Ngene, spdesign and idefix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading designs from Ngene, spdesign and idefix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(simulateDCE)
```

The other vignettes build small designs by hand so they run quickly. In practice
your design comes out of design software. `readdesign()` turns any of the usual
formats into the one data frame the rest of the package works with, and
`sim_all()` calls it for you on every file in a folder.

This vignette uses the real designs bundled with the package.

## What a design looks like once it is read

Whatever the source, you get one row per choice situation, a `Choice.situation`
column, and a `Block` column when the design is blocked:

```{r read-ngene}
ngene_file <- system.file("extdata", "SE_DRIVE", "effconstrsmall.ngd",
                          package = "simulateDCE")

design <- readdesign(ngene_file, designtype = "ngene")

dim(design)
names(design)
head(design[, 1:6], 4)
```

```{r blocks}
table(design$Block)
```

Sixty choice situations in five blocks, so each respondent sees 12.

## The four formats

### Ngene

Save the standard Ngene output with a `.ngd` extension. It is tab separated text,
and `readdesign()` drops the `Design` column and any trailing empty columns Ngene
writes.

```{r ngene2}
d <- readdesign(
  system.file("extdata", "agora", "altscf_eff.ngd", package = "simulateDCE"),
  designtype = "ngene"
)
c(situations = nrow(d), blocks = length(unique(d$Block)))
```

### spdesign

Save the whole object returned by `spdesign::generate_design()` with `saveRDS()`.
`readdesign()` takes the `design` element out of it:

```{r spdesign}
sp_file <- system.file("extdata", "ValuGaps", "des1.RDS", package = "simulateDCE")
class(readRDS(sp_file))

d <- readdesign(sp_file, designtype = "spdesign")
head(d, 3)
```

If your design uses underscores in the attribute names, the first underscore
becomes a dot, so `alt1_price` is read as `alt1.price`. That is the convention the
utility functions expect.

The priors you gave spdesign are stored in the object too, and
`extract_b_values()` pulls them out in the form `bcoeff` wants:

```{r priors}
csa <- readRDS(system.file("extdata", "CSA", "linear", "BLIeff.RDS",
                           package = "simulateDCE"))
extract_b_values(csa$utility)
```

That is a convenient starting point for a simulation: simulate from the priors the
design was optimised for, and you learn whether the design delivers on its own
assumptions.

### idefix

Save the object returned by `idefix::Modfed()` or `idefix::CEA()` with `saveRDS()`.
These store the design in long format with row names like `set1.alt1`, so
`readdesign()` reshapes it to one row per choice situation and drops the
`no.choice` rows and the `cte` columns:

```{r idefix}
d <- readdesign(
  system.file("extdata", "Idefix_designs", "test_design2.RDS", package = "simulateDCE"),
  designtype = "idefix"
)
dim(d)
names(d)[1:5]
```

### A plain data frame

If you built the design yourself, or exported it from something else, save it as a
data frame with one row per choice situation and use `designtype = "matrix"`. This
is what the other vignettes do.

## Letting it guess

Leave `designtype` out and the type is detected from the file:

```{r guess}
readdesign(ngene_file)                       |> dim()
readdesign(sp_file)                          |> dim()
```

The rules are:

| The file | Detected as |
|---|---|
| ends in `.ngd` | `ngene` |
| an RDS whose object has class `spdesign` | `spdesign` |
| an RDS holding a list with `design` plus `error`, `probs` or `set` row names | `idefix` |
| an RDS holding a plain data frame | `matrix` |
| an RDS holding a list with a `design` element and nothing idefix-like | `spdesign` |

Every rule tests for something positive. Anything it cannot place produces an
error naming what it found, rather than a guess:

```{r guess-fail, error = TRUE}
odd <- file.path(tempdir(), "odd.rds")
saveRDS(list(a = 1, b = 2), odd)
readdesign(odd)
```

Guessing is convenient when a folder mixes formats, since `sim_all()` reads
everything in it. Naming the type explicitly is still better when you know it: it
turns a wrong guess into an error rather than into a strange result.

## Running over a folder

`sim_all()` reads every matching file in `designpath` and simulates each one, so
you can compare designs directly:

```{r sim-all}
designpath <- system.file("extdata", "SE_DRIVE", package = "simulateDCE")
list.files(designpath)

bcoeff <- list(bpreis = -0.01, blade = -0.07, bwarte = 0.02)

ul <- list(
  u1 = list(
    v1 = V.1 ~ bpreis * alt1.x1 + blade * alt1.x2 + bwarte * alt1.x3,
    v2 = V.2 ~ bpreis * alt2.x1 + blade * alt2.x2 + bwarte * alt2.x3
  )
)

result <- sim_all(
  nosim = 20, resps = 200, designpath = designpath,
  u = ul, bcoeff = bcoeff, mode = "sequential", verbose = 0
)

sa <- result$summaryall
coef_rows <- !grepl("^rob_pval0_", sa$parname)

round(
  data.frame(
    truepar    = sa$truepar[coef_rows],
    small_mean = sa$effconstrsmall.mean[coef_rows],
    small_sd   = sa$effconstrsmall.sd[coef_rows],
    bay_mean   = sa$bayeffdesignconstr.mean[coef_rows],
    bay_sd     = sa$bayeffdesignconstr.sd[coef_rows],
    row.names  = sa$parname[coef_rows]
  ),
  4
)
```

Both designs recover the coefficients. The `sd` columns are the comparison you
came for: they say how precisely each design pins the parameters down at this
sample size. Power says the same thing in the units a reviewer will ask about:

```{r power}
result$powa_by_par
```

Only files matching `pattern` are treated as designs, by default anything ending
in `.ngd` or `.rds`. A stray note or plot in the folder is ignored rather than
breaking the run. Widen `pattern` if your designs use a different extension.

## Things worth checking

**Are the blocks equal?** Every respondent must see the same number of choice
situations, so blocks of different sizes are an error. Check with
`table(design$Block)`.

**Is the model identified?** With two alternatives, only the differences between
them carry information:

```{r identification}
d <- readdesign(ngene_file, designtype = "ngene", verbose = 0)

diffs <- with(d, data.frame(
  x1 = alt1.x1 - alt2.x1,
  x2 = alt1.x2 - alt2.x2,
  x3 = alt1.x3 - alt2.x3
))

c(
  rank        = qr(as.matrix(diffs))$rank,
  attributes  = ncol(diffs),
  largest_cor = round(max(abs(cor(diffs)[upper.tri(cor(diffs))])), 3)
)
```

Rank equal to the number of attributes means all of them are identified. This is a
sensible thing to run on any design before you spend a night simulating with it,
including one that came from design software.
