---
title: "Availability, mixed logit, and fitting your own model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Availability, mixed logit, and fitting your own model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(simulateDCE)
set.seed(3141)
```

Three things about the estimation half of the package: how to say that an
alternative was not always on offer, how to fit a mixed logit rather than a
multinomial one, and how to put a model of your own in the loop.

## A design with an opt-out

```{r design}
set.seed(3141)

profiles <- expand.grid(price = c(2, 4, 6, 8), quality = c(0, 1))
alt1 <- profiles[rep(seq_len(nrow(profiles)), 2), ]
n <- nrow(alt1)
repeat {
  partner <- sample(n)
  if (all(rowSums(abs(alt1 - alt1[partner, ])) > 0)) break
}

design <- data.frame(
  Choice.situation = seq_len(n),
  Block            = rep(1:2, each = n / 2),
  alt1.price = alt1$price, alt1.quality = alt1$quality,
  alt2.price = alt1$price[partner], alt2.quality = alt1$quality[partner],
  # the opt-out is offered in every other choice situation
  av1 = 1, av2 = 1, av3 = rep(c(1, 0), length.out = n)
)

design_dir <- file.path(tempdir(), "est_designs")
dir.create(design_dir, showWarnings = FALSE)
saveRDS(design, file.path(design_dir, "d.rds"))
```

A no-choice, opt-out or status-quo option is not a special kind of thing. It is an
alternative whose utility has no attributes, only a constant:

```{r utility}
ul <- list(u1 = list(
  v1 = V.1 ~ bprice * alt1.price + bquality * alt1.quality,
  v2 = V.2 ~ bprice * alt2.price + bquality * alt2.quality,
  v3 = V.3 ~ bnone
))

bcoeff <- list(bprice = -0.3, bquality = 0.5, bnone = -1.5)
```

`bnone` is the utility of choosing nothing, measured against the normalisation
that the other two alternatives carry no constant of their own.

## Availability

Columns named `av1`, `av2` and so on, holding 1 where the alternative was offered
and 0 where it was not, are picked up automatically. `av.1` and `av_1` work too.
With no such columns everything is available everywhere.

Availability matters in both halves of the package. When simulating, an
unavailable alternative cannot be chosen however high its utility comes out:

```{r simulate}
sim <- sim_all(
  nosim = 1, resps = 300, designpath = design_dir,
  u = ul, bcoeff = bcoeff, estimate = FALSE,
  mode = "sequential", seed = 1, verbose = 0
)$d[[1]]

table(offered = sim$av3, chosen = sim$CHOICE)
```

The zero in the top right is the point: the opt-out is never chosen in the choice
situations where it was withheld.

When estimating, the availability matrix goes to the model instead of a matrix of
ones, so the choice probabilities are normalised over what was actually on offer:

```{r estimate}
res <- sim_all(
  nosim = 20, resps = 400, designpath = design_dir,
  u = ul, bcoeff = bcoeff, mode = "sequential", seed = 2, verbose = 0
)

sa <- res$summaryall
keep <- !grepl("^rob_pval0_", sa$parname)
round(
  data.frame(
    truepar = sa$truepar[keep], estimate = sa$d.mean[keep],
    row.names = sa$parname[keep]
  ),
  3
)
```

`bnone` is recovered even though the opt-out was on offer only half the time. That
is the check worth making: if the availability had been ignored, the model would
have seen the opt-out chosen half as often relative to what it thought was
available, and put `bnone` too low.

## Mixed logit

By default the package fits a multinomial logit. Given heterogeneous preferences
that recovers roughly the mean of each coefficient's distribution, attenuated
towards zero, and says nothing about the spread. See
`vignette("random-parameters")` for how much that costs you.

`model = "mixed"` fits the model that matches the data generating process
instead. Each random entry of `bcoeff` gets its own respondent-level draw, so the
location and the spread are both estimated.

```{r mixed}
bcoeff_random <- list(
  bprice   = list(dist = "normal", mean = -0.4, sd = 0.25),
  bquality = 0.5,
  bnone    = -1.5
)

compare <- lapply(c("mnl", "mixed"), function(m) {
  res <- sim_all(
    nosim = 10, resps = 400, designpath = design_dir,
    u = ul, bcoeff = bcoeff_random, model = m, n_draws = 200,
    mode = "sequential", seed = 5, verbose = 0
  )
  sa <- res$summaryall
  stats::setNames(sa$d.mean, sa$parname)
})
names(compare) <- c("mnl", "mixed")

pars <- c("bprice", "sigma_bprice", "bquality", "bnone")
round(
  data.frame(
    truepar = c(-0.4, 0.25, 0.5, -1.5),
    mnl     = compare$mnl[pars],
    mixed   = compare$mixed[pars],
    row.names = pars
  ),
  3
)
```

The multinomial logit has no `sigma_bprice` to report, so that cell is missing, and
its `bprice` is pulled towards zero. The mixed logit recovers both the mean and the
standard deviation of the price coefficient.

### What the parameters mean

A mixed logit estimates a distribution's own parameters, not its moments. For a
`normal` those coincide: `bprice` is the mean and `sigma_bprice` the standard
deviation. For a `lognormal` or `neg_lognormal` they do not: `bprice` estimates
`meanlog` and `sigma_bprice` estimates `sdlog`, both on the log scale.
`truepar` follows the same convention, so the comparison stays honest either way.

```{r lognormal}
res <- sim_all(
  nosim = 10, resps = 400, designpath = design_dir,
  u = ul,
  bcoeff = list(
    bprice   = list(dist = "neg_lognormal", meanlog = -1, sdlog = 0.4),
    bquality = 0.5, bnone = -1.5
  ),
  model = "mixed", n_draws = 200,
  mode = "sequential", seed = 8, verbose = 0
)

sa <- res$summaryall
round(
  data.frame(
    truepar = sa$truepar[1:2], estimate = sa$d.mean[1:2],
    row.names = sa$parname[1:2]
  ),
  3
)
```

Two practical notes. Only `normal`, `lognormal` and `neg_lognormal` can be written
as a mixed logit here; the bounded shapes cannot, and asking for them is an error
rather than a silent approximation. And the sign of a scale parameter is not
identified, because the likelihood sees only `sigma * draw` and the draws are
symmetric about zero, so the optimiser lands on either sign with the same fit. The
absolute value is reported, as every mixed logit package does.

A mixed logit is warm-started from the multinomial fit, so it costs two models per
run rather than one, on top of being a harder model. Budget accordingly.

## Fitting your own model

The estimation backend is `mixl`. If you want something else, pass a function to
`estimator` and keep the simulation, aggregation and power machinery around it.

An estimator takes the simulated `data` and a `spec` describing the model, and
returns four things: `coefficients`, `pvalues`, `converged` and `model`. The most
useful part of `spec` is `model_matrix()`, which hands you each coefficient's
regressor for each alternative, so interactions and transformations are already
worked out.

Here is a conditional logit written from scratch:

```{r own-estimator}
plain_mnl <- function(data, spec) {
  mm <- spec$model_matrix()
  x <- mm$x # observations x alternatives x coefficients
  av <- spec$availabilities
  chosen <- cbind(seq_len(nrow(data)), data$CHOICE)

  negll <- function(b) {
    v <- apply(x, c(1, 2), function(r) sum(r * b))
    v[av == 0] <- -Inf
    -sum(v[chosen] - log(rowSums(exp(v))))
  }

  fit <- stats::optim(rep(0, length(mm$terms)), negll,
    method = "BFGS", hessian = TRUE
  )
  se <- sqrt(diag(solve(fit$hessian)))

  list(
    coefficients = stats::setNames(fit$par, mm$terms),
    pvalues = stats::setNames(2 * stats::pnorm(-abs(fit$par / se)), mm$terms),
    converged = fit$convergence == 0,
    model = fit
  )
}
```

Twenty lines, and it plugs straight in:

```{r use-own}
own <- sim_all(
  nosim = 10, resps = 300, designpath = design_dir,
  u = ul, bcoeff = bcoeff, estimator = plain_mnl,
  mode = "sequential", seed = 7, verbose = 0
)

built_in <- sim_all(
  nosim = 10, resps = 300, designpath = design_dir,
  u = ul, bcoeff = bcoeff,
  mode = "sequential", seed = 7, verbose = 0
)

a <- built_in$summaryall
b <- own$summaryall
keep <- !grepl("^rob_pval0_", a$parname)

round(
  data.frame(
    truepar   = a$truepar[keep],
    mixl      = a$d.mean[keep],
    own       = b$d.mean[keep],
    difference = a$d.mean[keep] - b$d.mean[keep],
    row.names = a$parname[keep]
  ),
  6
)
```

On the same simulated data the two agree to within about a millionth, the rest
being the difference between two optimisers' stopping rules. That is a reassuring
thing to know about both of them, and writing a second estimator is a practical way
to check the first.

Everything downstream still works: power, convergence counts and the aggregated
comparison across designs all read the standardised result rather than anything
`mixl`-specific.

```{r own-power}
own$powa_by_par

own$convergence$d[c("runs", "converged", "failed")]
```

### What to watch out for

`keep_models = TRUE` returns whatever you put in `model`, untouched, in the
numbered slots of the results. With the default backend that is a `mixl` object
carrying its own copy of the data; with the estimator above it is what `optim()`
returned.

In parallel mode your function is sent to worker processes, so it has to be
self-contained: use `::` for packages, and remember that anything holding an
external pointer, such as a compiled model, does not survive the trip and has to
be rebuilt inside the function. That is exactly why the default backend recompiles
its model once per worker rather than shipping one across.

See `?estimators` for the full contract.
