---
title: "Manipulating attributes before the utilities are computed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manipulating attributes before the utilities are computed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(simulateDCE)
set.seed(1177)
```

Your design file holds whatever the design software produced. That is not always
the scale you want to simulate on. Ngene might have written waiting time in
minutes when your prior is per hour, or the utility you have in mind might need a
term that is not a column at all. The `manipulations` argument changes columns
after the design has been expanded to respondents and before any utility is
computed, without you touching the design file.

## A design to work with

```{r design}
set.seed(1177)

profiles <- expand.grid(cost = c(200, 400, 600), wait = c(30, 60, 90))
alt1 <- profiles[rep(seq_len(nrow(profiles)), 2), ]
n <- nrow(alt1)

repeat {
  partner <- sample(n)
  if (all(rowSums(abs(alt1 - alt1[partner, ])) > 0)) break
}

design <- data.frame(
  Choice.situation = seq_len(n),
  Block            = rep(1:2, each = n / 2),
  alt1.cost        = alt1$cost,
  alt1.wait        = alt1$wait,
  alt2.cost        = alt1$cost[partner],
  alt2.wait        = alt1$wait[partner]
)

design_dir <- file.path(tempdir(), "man_designs")
dir.create(design_dir, showWarnings = FALSE)
saveRDS(design, file.path(design_dir, "design.rds"))

head(design, 4)
```

Cost is in euros, waiting time in minutes.

## Rescaling

Suppose your priors are per hundred euros and per hour. Pass a named list of
expressions built with `rlang::expr()`. The name is the column to create or
replace, the expression says how.

```{r manip}
manipulations <- list(
  alt1.cost = rlang::expr(alt1.cost / 100),
  alt2.cost = rlang::expr(alt2.cost / 100),
  alt1.wait = rlang::expr(alt1.wait / 60),
  alt2.wait = rlang::expr(alt2.wait / 60)
)

ul <- list(
  u1 = list(
    v1 = V.1 ~ bcost * alt1.cost + bwait * alt1.wait,
    v2 = V.2 ~ bcost * alt2.cost + bwait * alt2.wait
  )
)

bcoeff <- list(bcost = -0.4, bwait = -0.5)
```

```{r run}
result <- sim_all(
  nosim = 30, resps = 200, designpath = design_dir,
  u = ul, bcoeff = bcoeff, manipulations = manipulations,
  mode = "sequential", verbose = 0
)

sa <- result$summaryall
round(
  data.frame(
    truepar  = sa$truepar,
    estimate = sa$design.mean,
    row.names = sa$parname
  )[1:2, ],
  3
)
```

The coefficients are recovered on the rescaled attributes, because the
manipulation ran before the utilities were computed. The simulated data carries the
rescaled values too:

```{r data}
sim_only <- sim_all(
  nosim = 1, resps = 20, designpath = design_dir,
  u = ul, bcoeff = bcoeff, manipulations = manipulations,
  estimate = FALSE, mode = "sequential", verbose = 0
)

head(sim_only$design[[1]][, c("ID", "Choice_situation", "alt1_cost", "alt1_wait")], 4)
```

Cost is in hundreds and waiting time in hours. Note that by this point the column
names have had their dots turned into underscores, because that is what the
estimation needs. Inside `manipulations` you still write the dotted names, since
manipulations run before that rename.

## Why not just edit the design file?

You could, and sometimes you should. `manipulations` earns its place when:

- **The design file is not yours to change.** It came from a colleague, or it is
  the published design for a study you are replicating.
- **You want to compare scalings.** Running the same design at several
  transformations is one line of code rather than several files.
- **The transformation is part of the question.** Whether respondents react to
  cost or to log cost is a research question, not a correction.

## Creating new columns

Manipulations are not limited to rewriting existing columns. Anything computable
from the data is fair game, which is how you get interaction terms into a
simulation:

```{r new-columns}
ul_int <- list(
  u1 = list(
    v1 = V.1 ~ bcost * alt1.cost + bwait * alt1.wait + bboth * alt1.both,
    v2 = V.2 ~ bcost * alt2.cost + bwait * alt2.wait + bboth * alt2.both
  )
)

manip_int <- c(manipulations, list(
  alt1.both = rlang::expr(alt1.cost * alt1.wait),
  alt2.both = rlang::expr(alt2.cost * alt2.wait)
))

res <- sim_all(
  nosim = 20, resps = 250, designpath = design_dir,
  u = ul_int, bcoeff = c(bcoeff, list(bboth = 0.1)),
  manipulations = manip_int,
  mode = "sequential", verbose = 0
)

sa <- res$summaryall
round(
  data.frame(
    truepar  = sa$truepar,
    estimate = sa$design.mean,
    row.names = sa$parname
  )[1:3, ],
  3
)
```

The interaction is created *after* cost and wait have been rescaled, because it
comes later in the list. Order matters, and the next section returns to that.

### A word on identification

Not every column you can compute is a column you can estimate. A term that is a
linear function of the ones already there leaves the model unidentified, and no
sample size fixes it. The classic trap is a dummy for "the cheaper alternative" in
a design where the two alternatives mirror each other: being cheaper then follows
deterministically from the cost difference, and every run fails.

The check is cheap. Build the matrix of attribute differences, including your new
column, and look at its rank:

```{r identification}
diffs <- with(design, data.frame(
  cost = alt1.cost - alt2.cost,
  wait = alt1.wait - alt2.wait,
  both = (alt1.cost / 100 * alt1.wait / 60) - (alt2.cost / 100 * alt2.wait / 60)
))

c(rank = qr(as.matrix(diffs))$rank, terms = ncol(diffs))
```

Rank 3 for 3 terms means the interaction adds information rather than repeating
what is already there. If the rank came out lower, `sim_all()` would tell you that
no model produced usable estimates and point you here.

## Non-linear transformations

A common question is whether a linear cost term is good enough. Simulate from a log
specification to find out what the world would look like if it were not:

```{r nonlinear}
manip_log <- list(
  alt1.cost = rlang::expr(log(alt1.cost)),
  alt2.cost = rlang::expr(log(alt2.cost)),
  alt1.wait = rlang::expr(alt1.wait / 60),
  alt2.wait = rlang::expr(alt2.wait / 60)
)

res_log <- sim_all(
  nosim = 20, resps = 200, designpath = design_dir,
  u = ul, bcoeff = list(bcost = -2, bwait = -0.5),
  manipulations = manip_log,
  mode = "sequential", verbose = 0
)

sa <- res_log$summaryall
round(
  data.frame(
    truepar  = sa$truepar[1:2],
    estimate = sa$design.mean[1:2],
    row.names = sa$parname[1:2]
  ),
  3
)
```

Because the manipulation is applied before the utility, `bcost` here is the
coefficient on log cost, and it is recovered. To study *misspecification* instead,
simulate with the log manipulation, set `estimate = FALSE`, and fit the linear model
to the resulting data yourself.

## Manipulations run per respondent

Internally the manipulations are applied grouped by respondent, so expressions that
aggregate over a respondent's own choice tasks do what you would hope:

```{r grouped}
manip_grouped <- list(
  alt1.cost_rel = rlang::expr(alt1.cost / mean(alt1.cost))
)

d <- sim_all(
  nosim = 1, resps = 10, designpath = design_dir,
  u = ul, bcoeff = bcoeff, manipulations = manip_grouped,
  estimate = FALSE, mode = "sequential", verbose = 0
)$design[[1]]

head(d[, c("ID", "alt1_cost", "alt1_cost_rel")], 5)
```

`mean(alt1.cost)` is the mean over that respondent's own choice sets, not over the
whole sample. That is what you want for anything reference-dependent, and worth
knowing before you write an expression that assumes otherwise.

## Order matters

Manipulations are applied in the order you list them, and later ones see the
results of earlier ones:

```{r order}
sequential_manip <- list(
  alt1.cost    = rlang::expr(alt1.cost / 100),
  alt1.cost_sq = rlang::expr(alt1.cost^2) # squares the already divided value
)

d <- sim_all(
  nosim = 1, resps = 5, designpath = design_dir,
  u = ul, bcoeff = bcoeff, manipulations = sequential_manip,
  estimate = FALSE, mode = "sequential", verbose = 0
)$design[[1]]

unique(d[, c("alt1_cost", "alt1_cost_sq")])
```

`alt1_cost_sq` is the square of 2, 4 and 6, not of 200, 400 and 600. If you want
the square of the original, compute it before you rescale.
