---
title: "Random parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(simulateDCE)
set.seed(9081)
```

Assuming that every respondent shares the same coefficient is convenient and
usually wrong. People differ in how much they care about price. In
`simulateDCE` you say so by giving a coefficient a distribution instead of a
number, and each simulated respondent draws their own value from it.

This changes the data generating process, not the model you estimate. The package
still fits a multinomial logit to the simulated data, so this is the tool for
asking what happens to your MNL results when the truth is heterogeneous.

## From a number to a distribution

A fixed coefficient is a number. A random one is a list with a `dist` element and
the arguments that distribution needs:

```{r spec}
design <- data.frame(
  Choice.situation = 1:12,
  Block            = rep(1:2, each = 6),
  alt1.price       = rep(c(2, 4, 6), 4),
  alt1.quality     = rep(c(0, 1), 6),
  alt2.price       = rep(c(6, 4, 2), 4),
  alt2.quality     = rep(c(1, 0), 6)
)

design_dir <- file.path(tempdir(), "rp_designs")
dir.create(design_dir, showWarnings = FALSE)
saveRDS(design, file.path(design_dir, "design.rds"))

ul <- list(
  u1 = list(
    v1 = V.1 ~ bprice * alt1.price + bquality * alt1.quality,
    v2 = V.2 ~ bprice * alt2.price + bquality * alt2.quality
  )
)

bcoeff <- list(
  bprice   = list(dist = "normal", mean = -0.3, sd = 0.15),
  bquality = 0.4
)
```

Fixed and random coefficients mix freely. Here price varies across respondents
and quality does not.

## The distributions on offer

| `dist` | Arguments | Support |
|---|---|---|
| `"normal"` | `mean`, `sd` | unbounded |
| `"lognormal"` | `meanlog`, `sdlog` | strictly positive |
| `"neg_lognormal"` | `meanlog`, `sdlog` | strictly negative |
| `"uniform"` | `min`, `max` | bounded |
| `"triangular"` | `min`, `max`, `mode` | bounded, peaked |
| `"truncated_normal"` | `mean`, `sd`, `min`, `max` | bounded, either side may be infinite |
| `"fixed"` | `value` | a constant; the same as giving a bare number |

The bounded ones exist because sign matters. A normal price coefficient gives
some respondents a positive one, meaning they prefer paying more. If that is not
part of your story, use `"neg_lognormal"` or a `"truncated_normal"` capped at
zero.

### Watch the parameterisation

`lognormal` and `neg_lognormal` take `meanlog` and `sdlog`, which describe the
*underlying normal*, not the coefficient. `meanlog = 0` does not mean the
coefficient averages zero. `bcoeff_moments()` translates any specification into
the units of the coefficient itself:

```{r moments}
bcoeff_moments(list(
  bnormal    = list(dist = "normal", mean = -0.3, sd = 0.15),
  blognormal = list(dist = "lognormal", meanlog = 0, sdlog = 0.5),
  bneglog    = list(dist = "neg_lognormal", meanlog = -1, sdlog = 0.4),
  btriangle  = list(dist = "triangular", min = 0, max = 0.9, mode = 0.2),
  btruncated = list(dist = "truncated_normal", mean = 0, sd = 0.3,
                    min = -Inf, max = 0),
  bfixed     = 0.4
))
```

Run this before a long simulation. It is the quickest way to notice that the
distribution you specified is not the one you had in mind.

## Inspecting the draws

`make_rand_params()` gives you the respondent-level draws on their own, without
running a simulation:

```{r draws}
draws <- make_rand_params(bcoeff, n_resp = 500)
head(draws, 3)

round(c(mean = mean(draws$bprice), sd = sd(draws$bprice)), 3)

# what share of respondents ended up with a positive price coefficient?
mean(draws$bprice > 0)
```

A small share of respondents prefer paying more, because a normal distribution
centred at $-0.3$ with a standard deviation of $0.15$ puts about 2% of its mass
above zero. Whether that is acceptable is a modelling decision, and now it is one
you can see rather than one you made by accident.

## Running the simulation

Nothing else changes:

```{r run}
result <- sim_all(
  nosim      = 40,
  resps      = 300,
  designpath = design_dir,
  u          = ul,
  bcoeff     = bcoeff,
  mode       = "sequential"
)
```

The parameter summary now reports the distribution and its implied mean and
standard deviation for the random coefficient, and a plain value for the fixed
one.

## What an MNL recovers from heterogeneous data

The simulated respondents have their own price coefficients. The estimated model
has one. What does that single number end up being?

```{r recovery}
sa <- result$summaryall
coef_rows <- sa[sa$quantity == "estimate", ]

round(
  data.frame(
    truepar  = coef_rows$truepar,
    truesd   = coef_rows$truesd,
    estimate = coef_rows$design.mean,
    sd       = coef_rows$design.sd,
    row.names = coef_rows$parname
  ),
  3
)
```

Two columns describe the truth now. `truepar` is the mean of the mixing
distribution and `truesd` is its spread, which is 0 for a fixed coefficient.

The estimate lands close to `truepar` but slightly nearer zero. That attenuation
is expected rather than a defect, and it grows with the heterogeneity you ignore.
Holding the mean at $-0.3$ and raising only its spread:

```{r attenuation}
attenuation <- vapply(c(0, 0.15, 0.3, 0.6), function(s) {
  set.seed(7)
  spec <- if (s == 0) -0.3 else list(dist = "normal", mean = -0.3, sd = s)
  res <- sim_all(
    nosim = 40, resps = 400, designpath = design_dir, u = ul,
    bcoeff = list(bprice = spec, bquality = 0.4),
    mode = "sequential", verbose = 0
  )
  sa <- res$summaryall
  price <- sa$design.mean[sa$parname == "bprice"]
  qual  <- sa$design.mean[sa$parname == "bquality"]
  c(bprice = price, bquality = qual, wtp_quality = qual / -price)
}, numeric(3))

colnames(attenuation) <- paste0("sd = ", c(0, 0.15, 0.3, 0.6))
round(attenuation, 3)
```

Read the columns left to right. With no heterogeneity both coefficients are
recovered. As the spread of the price coefficient grows, ignoring it inflates the
unexplained part of utility and every coefficient shrinks towards zero, price most
of all because price is the one that varies.

The bottom row is the practical consequence. Willingness to pay for quality is the
ratio of the two coefficients, and because they shrink by different amounts the
ratio does not survive either: it drifts upward as heterogeneity rises. So this is
not a nuisance you can divide away. If preferences are heterogeneous and you fit an
MNL, both your coefficients and your WTP estimates are off, and this table is how
you find out by how much before it matters.

An MNL reports nothing at all about `truesd`. That is the information you give up
by not estimating a mixed logit, and seeing the size of what you give up is the
point of this exercise.

`sd` in the earlier table is a third quantity again: how much the *estimate* moves
between runs. Do not read it as heterogeneity.

## Which distribution matters

The same mean under different distributions does not give the same data. Here are
three price specifications with a mean near $-0.3$:

```{r compare}
specs <- list(
  normal    = list(dist = "normal", mean = -0.3, sd = 0.15),
  neg_log   = list(dist = "neg_lognormal", meanlog = -1.31, sdlog = 0.45),
  uniform   = list(dist = "uniform", min = -0.6, max = 0)
)

round(vapply(specs, function(s) {
  set.seed(4)
  res <- sim_all(
    nosim = 30, resps = 300, designpath = design_dir, u = ul,
    bcoeff = list(bprice = s, bquality = 0.4),
    mode = "sequential", verbose = 0
  )
  sa <- res$summaryall
  c(
    true_mean   = bcoeff_moments(list(b = s))$mean,
    estimate    = sa$design.mean[sa$parname == "bprice"],
    sd_of_est   = sa$design.sd[sa$parname == "bprice"]
  )
}, numeric(3)), 3)
```

All three recover close to their own mean, which is the reassuring result. The
spread of the estimates across runs differs, though, because a bounded
distribution produces less extreme respondents than an unbounded one.

## What is in the simulated data

With random parameters, each respondent's draw appears as a column, so you can
check the data generating process directly:

```{r data}
sim_only <- sim_all(
  nosim = 1, resps = 100, designpath = design_dir, u = ul,
  bcoeff = bcoeff, estimate = FALSE, mode = "sequential", verbose = 0
)

d <- sim_only$design[[1]]
head(d[, c("ID", "task", "bprice", "bquality", "V_1", "V_2", "CHOICE")], 4)

# one draw per respondent, constant across their choice tasks
all(tapply(d$bprice, d$ID, function(x) length(unique(x))) == 1)
```

The draw is redrawn for every simulation run, so the heterogeneity is part of
what is being simulated rather than a fixed feature of one dataset.

## Correlated preferences

By default the draws are independent across parameters, so a respondent who is
price sensitive is not systematically also quality sensitive. Real preferences are
rarely like that. Pass a `correlation` matrix and they need not be:

```{r correlated}
bcoeff_corr <- list(
  bprice = list(dist = "normal", mean = -0.3, sd = 0.15),
  bqual  = list(dist = "normal", mean = 0.4, sd = 0.2)
)

r <- correlate(c("bprice", "bqual") ~ 0.6)
r
```

```{r correlated-draws}
set.seed(20)
draws <- make_rand_params(bcoeff_corr, n_resp = 5000, correlation = r)

round(c(
  requested = 0.6,
  realised  = stats::cor(draws$bprice, draws$bqual),
  mean_price = mean(draws$bprice),
  sd_price   = stats::sd(draws$bprice)
), 3)
```

The correlation comes out where you asked, and the marginals are untouched.

Under the hood this is a Gaussian copula: correlated standard normals are drawn,
turned into uniforms, and pushed through each coefficient's own quantile function.
That is what lets a lognormal price coefficient be correlated with a triangular
quality coefficient, which no simpler scheme manages:

```{r correlated-mixed}
set.seed(21)
mixed_marginals <- make_rand_params(
  list(
    bprice = list(dist = "neg_lognormal", meanlog = -1, sdlog = 0.4),
    bqual  = list(dist = "triangular", min = 0, max = 1, mode = 0.3)
  ),
  n_resp = 5000,
  correlation = correlate(c("bprice", "bqual") ~ 0.6)
)

round(c(
  pearson  = stats::cor(mixed_marginals$bprice, mixed_marginals$bqual),
  spearman = stats::cor(mixed_marginals$bprice, mixed_marginals$bqual,
                        method = "spearman"),
  all_price_negative = all(mixed_marginals$bprice < 0),
  qual_within_bounds = all(mixed_marginals$bqual >= 0 & mixed_marginals$bqual <= 1)
), 3)
```

The number you give is the correlation of the underlying normals. With normal
marginals that is exactly the correlation of the coefficients. With other marginals
the rank correlation is preserved exactly and the Pearson correlation comes out
close, because the mapping to the marginal is not linear. Both are reported above
so you can see the difference.

Pass `correlation` to `sim_all()` the same way, and the respondents in the
simulated data have correlated tastes.

## Notes and limits

Correlation enters the **data generating process**. Estimating it is another
matter: `model = "mixed"` fits independent random parameters whatever
`correlation` says. That is not an oversight so much as the interesting case, since
the gap between the two is exactly what you would be measuring.

Draws are **pseudo-random**, not Halton or MLHS. That is deliberate: quasi-random
sequences belong in the *estimation* of a mixed logit, where they approximate an
integral. Here you are generating a population, and for that you want genuine
randomness.

The **model estimated is still an MNL**. To fit a mixed logit to these data, set
`estimate = FALSE`, take the simulated datasets, and pass them to `mixl` or
`apollo` yourself.
