---
title: "Working with the results, and running big simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with the results, and running big simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(simulateDCE)
set.seed(4004)
```

`sim_all()` returns a large nested list. This vignette says what is in it, and then
covers the arguments that matter once a simulation stops being a five second
demonstration: parallel processing, memory, saving partial results, and merging
runs done at different times.

## A run to look at

```{r run}
set.seed(4004)

profiles <- expand.grid(price = c(2, 4, 6, 8), quality = c(0, 1))
alt1 <- profiles[rep(seq_len(nrow(profiles)), 2), ]
n <- nrow(alt1)
repeat {
  partner <- sample(n)
  if (all(rowSums(abs(alt1 - alt1[partner, ])) > 0)) break
}

design <- data.frame(
  Choice.situation = seq_len(n),
  Block            = rep(1:2, each = n / 2),
  alt1.price = alt1$price, alt1.quality = alt1$quality,
  alt2.price = alt1$price[partner], alt2.quality = alt1$quality[partner]
)

design_dir <- file.path(tempdir(), "res_designs")
dir.create(design_dir, showWarnings = FALSE)
saveRDS(design, file.path(design_dir, "designA.rds"))
saveRDS(design[nrow(design):1, ], file.path(design_dir, "designB.rds"))

ul <- list(u1 = list(
  v1 = V.1 ~ bprice * alt1.price + bquality * alt1.quality,
  v2 = V.2 ~ bprice * alt2.price + bquality * alt2.quality
))
bcoeff <- list(bprice = -0.25, bquality = 0.4)

result <- sim_all(
  nosim = 30, resps = 200, designpath = design_dir,
  u = ul, bcoeff = bcoeff, mode = "sequential", verbose = 0
)
```

## What is in the result

```{r structure}
names(result)
```

The first entries are named after your design files. Everything after them is
about the run as a whole. (`arguements` is a misspelling that used to be the only
name for `arguments`; both are there so old scripts keep working.)

### Per design

`result[["designA"]]` holds that design's own results. The estimated models sit in
unnamed slots, so the named ones are the summaries:

```{r per-design}
nm <- names(result$designA)
nm[nzchar(nm)]

# and one unnamed slot per run
sum(!nzchar(nm))
```

- **`coefs`** one row per run, with an `est_` column per coefficient and a
  `rob_pval0_` column per robust p value.
- **`summary`** those columns summarised over the runs.
- **`power`** the share of runs in which every coefficient was significant.
- **`power_by_par`** the same per coefficient.
- **`convergence`** how many models converged, and why the rest did not.
- **`metainfo`**, **`bcoeff`**, **`designname`** what the run was, so a saved file
  describes itself.

The unnamed slots are the estimated model objects, one per run:

```{r models}
class(result$designA[[1]])

# the first run's simulated data is inside its model
head(result$designA[[1]]$data[, c("ID", "task", "CHOICE")], 3)
```

### Across designs

```{r across}
head(result$estimates, 3)
```

`estimates` is the long table behind the plots: one row per run per design, with a
`design` column. It is the thing to use for any comparison you want to make
yourself.

```{r summaryall}
sa <- result$summaryall
coef_rows <- !grepl("^rob_pval0_", sa$parname)

round(
  data.frame(
    truepar = sa$truepar[coef_rows],
    A_mean  = sa$designA.mean[coef_rows],
    A_sd    = sa$designA.sd[coef_rows],
    B_mean  = sa$designB.mean[coef_rows],
    B_sd    = sa$designB.sd[coef_rows],
    row.names = sa$parname[coef_rows]
  ),
  4
)
```

`summaryall` also carries a `median` column per design. If the mean and the median
disagree for a coefficient, one run has gone somewhere strange, and `min`, `max` and
`range` will show you where.

`graphs` holds one density plot per coefficient, comparing the designs. They are
returned rather than drawn, so you decide where they go:

```{r graphs, fig.width = 6, fig.height = 3.5}
names(result$graphs)

result$graphs$bprice
```

## Convergence

Worth a glance on every real run:

```{r convergence}
result$convergence$designA
```

A model that does not converge is left out of the summaries, and `sim_all()` warns
when that happens. Non-finite estimates count as failures too, because a separated
logit can report success while a coefficient has run off to several hundred. If
`failed` is more than a handful, the usual cause is a design that cannot identify
the model rather than anything about the sample size.

## Simulating without estimating

```{r no-estimate}
datasets <- sim_all(
  nosim = 3, resps = 100, designpath = design_dir,
  u = ul, bcoeff = bcoeff, estimate = FALSE,
  mode = "sequential", verbose = 0
)

length(datasets$designA)
names(datasets$designA[[1]])
```

Each run is a data frame carrying the design, the deterministic utilities `V_k`,
the Gumbel draws `e_k`, their sums `U_k`, and `CHOICE`. Use this when you want to
estimate with `apollo`, a mixed logit, or anything else the package does not fit
itself.

## Running something big

### Parallel processing

`mode = "parallel"` spreads the runs over several R processes. Starting eight of
them and shipping the data across costs on the order of ten seconds, which has to
be earned back. On the small three-coefficient model used above:

| Runs | Respondents | Sequential | 8 workers | Speedup |
|---|---|---|---|---|
| 100 | 300 | 6.5 s | 12.6 s | 0.5 |
| 200 | 300 | 12.8 s | 15.0 s | 0.9 |
| 200 | 1000 | 29.3 s | 21.7 s | 1.4 |
| 500 | 300 | 32.0 s | 20.4 s | 1.6 |

The rule of thumb is to use parallel when a sequential run would take more than
about half a minute, and not before. Bigger models shift the balance further
towards parallel, because the startup cost stays the same while the work grows, so
a mixed logit or a model with a dozen coefficients benefits much sooner than this
table suggests. Time your own case rather than assuming.

One thing to know about reproducibility. Both modes reproduce from a seed, but they
do not produce the *same* numbers, because parallel runs draw from separate
L'Ecuyer streams rather than from this session's stream. What parallel results do
not depend on is how many workers you use, so a rerun on a different allocation
still reproduces. So set a seed and stick to one mode; do not expect a sequential
run and a parallel run to agree run for run.

`workers` sets the number of processes. Leave it out and `future` picks, which on a
cluster means it reads `SLURM_CPUS_PER_TASK` and does the right thing. If you have
already set up a `future` plan yourself, `sim_all()` uses it and hands it back
untouched.

### Memory

Every estimated model carries its own copy of the simulated data, and with
thousands of runs that is what fills the memory rather than anything else. Sixty
runs of 300 respondents came to 23 MB with the models kept and 0.08 MB without.
Dropping the models is what does the work; dropping the utility columns on its own
only saved about a quarter.

```{r lean}
lean <- sim_all(
  nosim = 20, resps = 200, designpath = design_dir,
  u = ul, bcoeff = bcoeff,
  keep_models = FALSE, keep_utilities = FALSE,
  mode = "sequential", verbose = 0
)

c(
  full_MB = round(as.numeric(object.size(result)) / 1024^2, 1),
  lean_MB = round(as.numeric(object.size(lean)) / 1024^2, 1)
)

# everything the comparison needs is still there
names(lean$designA)
```

`keep_models = FALSE` drops the model objects and keeps the coefficients, summary,
power and convergence results, which is all `aggregateResults()` uses.
`keep_utilities = FALSE` drops the `V_`, `e_` and `U_` columns from the simulated
data, which the estimation does not need either. Use both for long runs, and leave
them on when you want to look at the data.

### Writing results to disk as you go

Two arguments guard against losing a long run.

`chunks` writes what has been estimated so far to a scratch file every so often.
With `nosim = 1000` and `chunks = 10` that happens every 100 runs. The scratch
files live in the session's temporary directory and are cleaned up at the end.

`savefile` writes one file per design as each design finishes:

```{r savefile}
save_dir <- file.path(tempdir(), "saved")

invisible(sim_all(
  nosim = 20, resps = 150, designpath = design_dir,
  u = ul, bcoeff = bcoeff, savefile = file.path(save_dir, "run"),
  mode = "sequential", verbose = 0
))

list.files(save_dir)
```

## Merging runs from different times

This is what `savefile` is really for. Simulate three designs today, add a fourth
next month, and put the results together:

```{r merge}
# a fourth design arrives later
later <- design
later$alt1.price <- later$alt1.price + 1
saveRDS(later, file.path(design_dir, "designC.rds"))

invisible(sim_all(
  nosim = 20, resps = 150, designpath = design_dir,
  pattern = "designC\\.rds$",
  u = ul, bcoeff = bcoeff, savefile = file.path(save_dir, "run"),
  mode = "sequential", verbose = 0
))

list.files(save_dir)

merged <- aggregateResults(fromfolder = save_dir)

unique(merged$estimates$design)
names(merged$powa)
```

Each saved file records its own `designname` and `bcoeff`, so the folder describes
itself and `aggregateResults()` needs nothing else. Note the `pattern` argument
above, which restricted the second run to the new design rather than redoing all
three.

## On a cluster

A few things that matter on SLURM specifically.

Set `--cpus-per-task` and leave `workers` alone. `future` reads
`SLURM_CPUS_PER_TASK` through `parallelly::availableCores()`, so the number of
processes matches what you were allocated.

Stop the linear algebra libraries from oversubscribing on top of that:

```
export OMP_NUM_THREADS=1
export OPENBLAS_NUM_THREADS=1
export MKL_NUM_THREADS=1
```

Set a seed in the script, not on the command line, so the run is reproducible from
the script alone. Parallel results do not depend on the worker count, so a rerun on
a different allocation still reproduces.

Use `keep_models = FALSE` and `savefile` for anything long. The memory limit that
usually bites is per core, not total, so a run that fits comfortably on your laptop
can fail on a node with more cores and less memory each.
