---
title: "Using Manipulations to Derive Variables at Simulation Time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Manipulations to Derive Variables at Simulation Time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!-- Build instructions
1. Create vignette using usethis::use_vignette("my-vignette")
2. After making changes, run devtools::build_vignettes()
3. Rebuild using devtools::install(build_vignettes = TRUE)
4. Check with browseVignettes()

If the vignette does not appear on GitHub, rendered files may have landed in /doc
instead of /inst/doc. To fix:
  tools::buildVignettes(dir = ".", tangle = TRUE)
  dir.create("inst/doc")
  file.copy(dir("vignettes", full.names = TRUE), "inst/doc", overwrite = TRUE)
-->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette introduces the `manipulations` argument in `sim_all()`. It builds
on the basic workflow described in the CSA vignette; if you haven't read that
one yet it is worth starting there.

The need for manipulations arises because experimental design software often
stores attribute levels as **integer codes** (e.g. 1, 2, 3) rather than the
dummy variables your utility function actually needs. Rather than pre-processing
the design file by hand, you can supply a named list of expressions to
`sim_all()` via the `manipulations` argument. These expressions are evaluated
inside the dataset *before* utility is calculated, so you can derive any new
variables on the fly.

The example here uses a Swedish agricultural subsidy study (`SE_AGRI`). The
design has two relevant attributes stored as integer codes:

- **`initiator`**: who initiates the subsidy scheme — `1` = professional
  organisation, `2` = expert body
- **`funding`**: source of funding — `1` = domestic, `2` = foreign

The utility function is written in terms of dummy variables
(`professional`, `expert`, `domestic`, `foreign`), so we need to create
those dummies at simulation time using manipulations.

```{r setup}
library(simulateDCE)
library(rlang)
library(formula.tools)
```

---

# Step 1 — Specify manipulations, coefficients, and utility functions

## Manipulations

A manipulation is a named expression that is evaluated row-by-row in the
simulated dataset. The name on the left becomes a new column; the expression on
the right can reference any column that already exists in the design.

Here we create four dummy variables — two for `initiator` and two for `funding`
— for each alternative (`alt1` and `alt2`):

```{r initialize}
manipulations <- list(
  alt1.professional = expr(alt1.initiator == 1),
  alt2.professional = expr(alt2.initiator == 1),
  alt1.expert       = expr(alt1.initiator == 2),
  alt2.expert       = expr(alt2.initiator == 2),
  alt1.domestic     = expr(alt1.funding == 1),
  alt2.domestic     = expr(alt2.funding == 1),
  alt1.foreign      = expr(alt1.funding == 2),
  alt2.foreign      = expr(alt2.funding == 2)
)
```

Each `expr()` is a **quoted expression** (from the `rlang` package) that is
injected into a `dplyr::mutate()` call. The result is `TRUE`/`FALSE`, which R
treats as `1`/`0` in arithmetic — exactly what is needed for a dummy variable
in a linear utility formula.

> **Tip:** you can use any expression that `dplyr::mutate()` would accept —
> arithmetic rescaling (`expr(price / 100)`), log transforms, interaction
> terms, and so on.

## Beta coefficients

These are the hypothesised true preferences of the population. `basc` is the
alternative-specific constant for the opt-out alternative (v3), which captures
the baseline tendency to choose the status quo. The positive value here means
the opt-out is systematically preferred — a realistic assumption in subsidy
choice experiments.

```{r bcoeff}
bcoeff <- list(
  basc      =  4.2,  # strong preference for the opt-out (status quo)
  bprof     =  0.3,  # professional initiator
  bexp      =  0.3,  # expert initiator
  bdomestic =  0.3,  # domestically funded
  bforeign  =  0.3,  # foreign funded
  bdamage   =  0.6,  # damage level (higher = more preferred, e.g. less damage)
  bprice    =  0.2   # compensation level
)
```

## Utility functions

The utility functions reference the derived dummy columns by name. Because
`alt1.professional` is created by the manipulation above, it is available here
as if it had always been in the design file.

```{r utility}
ul <- list(
  u1 = list(
    v1 = V.1 ~ bprof * alt1.professional + bexp * alt1.expert +
               bdomestic * alt1.domestic + bforeign * alt1.foreign +
               bdamage * alt1.damage + bprice * alt1.compensation,
    v2 = V.2 ~ bprof * alt2.professional + bexp * alt2.expert +
               bdomestic * alt2.domestic + bforeign * alt2.foreign +
               bdamage * alt2.damage + bprice * alt2.compensation,
    v3 = V.3 ~ basc   # opt-out
  )
)
```

---

# Extension — Decision groups

Real survey respondents do not always follow the same decision process. Some
may ignore certain attributes entirely, or apply a simpler heuristic. The
`decisiongroups` argument lets you partition respondents into subgroups, each
with its own utility function.

## How it works

`decisiongroups` is a numeric vector of **cumulative proportions** that sums
to 1 and starts at 0. Each interval defines one group:

```
decisiongroups = c(0, 0.6, 1)   →   group 1: first 60 %, group 2: remaining 40 %
```

The `u` list must then contain one sub-list per group, in the same order.
The groups are applied to the respondents in the order they appear in the
dataset, so the split is deterministic given a fixed seed.

## Example: attribute non-attendance

A common real-world finding is that a proportion of respondents ignores the
*initiator* attribute entirely when making their choices — they only respond
to damage level and compensation. We can simulate this by giving that subgroup
a utility function that omits `bprof` and `bexp`.

```{r decision-groups}
# Group 1 (70 %): attends to all attributes, including initiator type
# Group 2 (30 %): ignores initiator — only responds to damage and compensation

ul_heuristic <- list(
  u1 = list(
    v1 = V.1 ~ bprof * alt1.professional + bexp * alt1.expert +
               bdomestic * alt1.domestic + bforeign * alt1.foreign +
               bdamage * alt1.damage + bprice * alt1.compensation,
    v2 = V.2 ~ bprof * alt2.professional + bexp * alt2.expert +
               bdomestic * alt2.domestic + bforeign * alt2.foreign +
               bdamage * alt2.damage + bprice * alt2.compensation,
    v3 = V.3 ~ basc
  ),
  u2 = list(
    v1 = V.1 ~ bdamage * alt1.damage + bprice * alt1.compensation,
    v2 = V.2 ~ bdamage * alt2.damage + bprice * alt2.compensation,
    v3 = V.3 ~ basc
  )
)

decisiongroups <- c(0, 0.7, 1)   # 70 % in group 1, 30 % in group 2
```

Now pass both `ul_heuristic` and `decisiongroups` to `sim_all()`. Everything
else — manipulations, bcoeff, design path — stays the same.

```{r run-heuristic, eval = FALSE}
set.seed(3393)

seagri_heuristic <- simulateDCE::sim_all(
  nosim          = nosim,
  resps          = resps,
  designtype     = destype,
  designpath     = designpath,
  u              = ul_heuristic,
  bcoeff         = bcoeff,
  manipulations  = manipulations,
  decisiongroups = decisiongroups
)
```

> **Note:** The `manipulations` are applied to the full dataset before any
> group split occurs, so derived dummy variables are available to all utility
> functions regardless of which group uses them.

## Interpreting the results

When decision groups are used, the estimated model is still fitted on the
*combined* dataset. The simulated choices reflect the mixture of behaviours,
so the estimated coefficients represent a weighted average across groups. This
is useful for assessing whether a standard MNL model can still recover
meaningful parameter estimates when a fraction of respondents uses a simplified
heuristic.

To investigate how the share of the non-attending group affects power, you can
run `sim_all()` multiple times with different splits (e.g. `c(0, 0.5, 1)`,
`c(0, 0.8, 1)`) and compare the resulting summary tables.

---

# Step 2 — Set the design path and simulation parameters

The three Bayesian-efficient designs for this study are stored in the
`SE_AGRI` folder that ships with the package. We run all of them in one call.

```{r other}
designpath <- system.file("extdata", "SE_AGRI", package = "simulateDCE")
# For your own designs, use the folder path directly, e.g.:
# designpath <- "Projects/SE_Agri/Designs/"

resps   <- 240  # number of respondents per simulated dataset
nosim   <- 2    # simulation runs (use >= 500 for real power analyses)
destype <- "ngene"
```

---

# Step 3 — Set a seed and run

```{r random}
set.seed(3393)
```

Pass `manipulations` to `sim_all()` alongside the other arguments. The
manipulations are applied inside `simulate_choices()` before any utility
is computed, so the derived columns are available throughout.

```{r output}
seagri <- simulateDCE::sim_all(
  nosim        = nosim,
  resps        = resps,
  designtype   = destype,
  designpath   = designpath,
  u            = ul,
  bcoeff       = bcoeff,
  manipulations = manipulations
)
```

---

# Step 4 — Inspect the results

The return value has the same structure as in the CSA example: a nested list
keyed by design name, with `coefs` (one row per simulation run) and `summary`
(statistics across all runs).

```{r accessOutput}
# Names of the designs that were processed
designs <- names(seagri[sapply(seagri, is.list)])
print(designs)

# Estimated coefficients and summary for the first design
simulationCoeff <- seagri[[1]]$coefs
coeffSummary    <- seagri[[1]]$summary

print(simulationCoeff)
print(coeffSummary)

# To save results to disk:
# saveRDS(seagri, file = "results/seagri_simulation.RDS")
```

The `summary` table shows mean estimated values and their spread across the
`nosim` replications. When `nosim` is large enough (≥ 500), the percentage
of replications where each coefficient is statistically significant gives you
the **power** of the design for detecting that parameter.

---

# Summary

This vignette covered two arguments that extend the basic simulation workflow:

**`manipulations`** — derive new columns from design attributes at simulation
time, without modifying the design file.

- Supply a named list of `rlang::expr()` expressions.
- Expressions are evaluated inside `dplyr::mutate()` before utility is computed.
- Any `mutate()`-compatible expression is valid: comparisons, arithmetic,
  log transforms, interactions, etc.

**`decisiongroups`** — partition respondents into subgroups, each following a
different utility function.

- Supply a numeric vector of cumulative proportions starting at 0 and ending
  at 1 (e.g. `c(0, 0.7, 1)` for a 70/30 split).
- The `u` list must contain one sub-list per group, in matching order.
- Manipulations are applied to all respondents before the group split, so
  derived variables are available to every group's utility function.
- Useful for simulating attribute non-attendance, lexicographic heuristics,
  or any situation where a proportion of respondents behaves differently.
