---
title: "Getting started with simulateDCE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with simulateDCE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(simulateDCE)
set.seed(5432)
```

You have a choice experiment design and a hypothesis about how people will
respond to it. Before you go to the field, you want to know whether that design
and your planned sample size can actually recover the effects you care about.
That is what this package is for: you write down the data generating process,
it simulates respondents who behave that way, estimates your model on the
simulated data many times over, and reports how the estimates behave.

This vignette walks through one complete simulation. Later vignettes cover
individual features:

| Vignette | Covers |
|---|---|
| `vignette("design-files")` | Reading designs from Ngene, spdesign and idefix |
| `vignette("random-parameters")` | Coefficients that vary across respondents |
| `vignette("decision-groups")` | Subgroups that follow different decision rules |
| `vignette("manipulations")` | Changing attributes before utilities are computed |
| `vignette("unblocked-designs")` | Drawing choice sets at random per respondent |
| `vignette("results")` | What comes back, and how to work with it |
| `vignette("estimators")` | Availability, mixed logit, and fitting your own model |

## A design to simulate from

Normally you would read a design produced by Ngene, spdesign or idefix. To keep
this vignette self-contained and quick, I build a small one by hand. It is an
ordinary data frame with one row per choice situation.

```{r design}
design <- data.frame(
  Choice.situation = 1:12,
  Block            = rep(1:2, each = 6),
  alt1.price       = rep(c(2, 4, 6), 4),
  alt1.quality     = rep(c(0, 1), 6),
  alt2.price       = rep(c(6, 4, 2), 4),
  alt2.quality     = rep(c(1, 0), 6)
)

design
```

Before simulating anything, check that the design can actually identify the model
you have in mind. A design whose attributes always move together will still
converge and still fill in a summary table; the numbers will just be wrong.

```{r check}
check_design(design)
```

Two conventions matter. `Choice.situation` numbers the rows. Attributes are
named `alt<k>.<attribute>`, so the utility functions can refer to them. The
`Block` column splits the 12 choice situations into 2 blocks of 6, so each
respondent answers 6 choice tasks rather than all 12.

`sim_all()` reads designs from a folder, so save it to one:

```{r folder}
design_dir <- file.path(tempdir(), "my_designs")
dir.create(design_dir, showWarnings = FALSE)
saveRDS(design, file.path(design_dir, "efficient.rds"))
```

## The data generating process

Two pieces describe how your simulated respondents behave.

**The utility functions.** One formula per alternative, wrapped in a list. The
outer list holds one entry per decision group, so with a single group you still
need the inner list. Name the left hand sides `V.1`, `V.2`, and so on, in order.

```{r utility}
ul <- list(
  u1 = list(
    v1 = V.1 ~ bprice * alt1.price + bquality * alt1.quality,
    v2 = V.2 ~ bprice * alt2.price + bquality * alt2.quality
  )
)
```

**The coefficients.** One entry per parameter in the utility functions. Names
must start with a lower case `b`, because that is how the package tells
coefficients apart from attributes.

```{r bcoeff}
bcoeff <- list(
  bprice   = -0.15,
  bquality =  0.15
)
```

That is the whole story: everyone weights price at $-0.15$ and quality at $0.15$,
adds a random draw from a standard Gumbel distribution to each alternative, and
picks whichever alternative comes out highest. This is exactly the multinomial
logit model, so it is a data generating process an MNL should be able to
recover. Price varies over a range of 4 and quality over a range of 1, so price
does far more work here than the equal coefficients suggest. That asymmetry turns
out to matter below.

## Running the simulation

```{r run}
result <- sim_all(
  nosim      = 40,
  resps      = 180,
  designpath = design_dir,
  u          = ul,
  bcoeff     = bcoeff,
  mode       = "sequential"
)
```

`nosim` is how many times the whole experiment is repeated. Forty is enough to
see something here; for a result you would report, use several hundred. `resps`
is the sample size you are testing.

The parameter summary at the top repeats what you asked for, which is worth
reading: it is the cheapest way to catch a coefficient you meant to set
differently. The summary table then reports, across the 30 runs, the mean of each
estimate, how much it varied, and its range. The power line says in what share of
runs every coefficient came out significant at 5%.

## Reading the result

The comparison table across designs lives in `summaryall`. It holds one row per
estimated coefficient and one per robust p value, so it is easier to read a
column at a time:

```{r summary}
sa <- result$summaryall
coef_rows <- sa[sa$quantity == "estimate", ]

round(
  data.frame(
    truepar  = coef_rows$truepar,
    estimate = coef_rows$efficient.mean,
    sd       = coef_rows$efficient.sd,
    min      = coef_rows$efficient.min,
    max      = coef_rows$efficient.max,
    row.names = coef_rows$parname
  ),
  3
)
```

`truepar` is the value you put in, and `estimate` is the average across the runs.
They should be close, and the difference tells you about bias. `sd` is how much
the estimate moves from one run to the next, which is the sampling variability you
would face with this design and this sample size. Note how much wider the range is
for quality than for price.

Power is reported per design:

```{r power}
result$powa

result$powa_by_par
```

`powa` is joint power: the share of runs in which *every* coefficient was
significant. `powa_by_par` breaks it down, which is usually what you want, since
one weak attribute drags joint power down while everything else is fine. That is
exactly what happens here, and it is the whole reason to look at both.

## Is the sample size enough?

The reason to simulate is to answer that question before collecting data. Run the
same design at several sample sizes and compare:

```{r power-curve}
curve <- power_curve(
  resps      = c(60, 120, 240, 480),
  nosim      = 40,
  designpath = design_dir,
  u          = ul,
  bcoeff     = bcoeff,
  mode       = "sequential",
  seed       = 11,
  verbose    = 0
)

curve[curve$parameter != "(all)", c("parameter", "resps", "power", "se")]
```

Price is already detected in nearly every run at 60 respondents. Quality needs
several hundred, because it varies over a range of 1 against price's range of 4.
Two attributes with identical coefficients can therefore need very different
sample sizes, and only the simulation tells you which. If your design has to
support a claim about quality, this table is the argument for the sample size you
ask a funder to pay for.

With 40 runs these percentages are themselves noisy, roughly plus or minus 8
points. Raise `nosim` before you rely on them.

## Where to go next

Set `estimate = FALSE` if you only want simulated datasets and intend to estimate
elsewhere, for instance with `apollo` or `mixl` directly:

```{r no-estimate}
sim_only <- sim_all(
  nosim = 2, resps = 60, designpath = design_dir,
  u = ul, bcoeff = bcoeff, estimate = FALSE,
  mode = "sequential", verbose = 0
)

head(sim_only$efficient[[1]], 4)
```

Each run is one data frame. `V_1` and `V_2` are the deterministic utilities,
`e_1` and `e_2` the Gumbel draws, `U_1` and `U_2` their sums, and `CHOICE` the
alternative with the highest total utility. Having the true utilities alongside
the simulated choice is what makes this useful for teaching.

From here, `vignette("random-parameters")` relaxes the assumption that everyone
shares the same coefficients, and `vignette("results")` goes through the rest of
the output.
