---
title: "Unblocked designs and random choice sets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unblocked designs and random choice sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(simulateDCE)
set.seed(6180)
```

A blocked design decides in advance which choice sets go together. Block 1 is
these eight tasks, block 2 is those eight, and every respondent gets one block.

Sometimes you do not want that. You have a design with 40 choice situations, you
want each respondent to answer 8, and you would rather draw those 8 at random for
each person than commit to five fixed blocks. Set `sets_per_resp` and the package
does that.

## The basic idea

```{r design}
set.seed(6180)

# 10 profiles, each used 4 times, gives 40 choice situations
profiles <- expand.grid(price = c(2, 4, 6, 8, 10), quality = c(0, 1))
alt1 <- profiles[rep(seq_len(nrow(profiles)), 4), ]
n <- nrow(alt1)

# pair each profile with a different one
repeat {
  partner <- sample(n)
  if (all(rowSums(abs(alt1 - alt1[partner, ])) > 0)) break
}

design <- data.frame(
  Choice.situation = seq_len(n),
  alt1.price       = alt1$price,
  alt1.quality     = alt1$quality,
  alt2.price       = alt1$price[partner],
  alt2.quality     = alt1$quality[partner]
)

# no Block column: blocks and random draws are alternatives, not partners
"Block" %in% names(design)

nrow(design)
```

```{r draw}
allocation <- draw_sets(design, respondents = 100, sets_per_resp = 8)

head(allocation[, c("ID", "task", "Choice.situation", "alt1.price")], 10)
```

Each respondent gets 8 rows. `task` numbers them 1 to 8 in the order they were
shown, and `Choice.situation` records which row of the design that was. Keeping
both matters: `task` is the choice occasion your panel model needs, and
`Choice.situation` is what tells you which set was actually seen.

```{r shape}
c(rows = nrow(allocation), respondents = length(unique(allocation$ID)))

# every respondent sees 8 distinct sets, never the same one twice
all(table(allocation$ID) == 8)
all(tapply(allocation$Choice.situation, allocation$ID, anyDuplicated) == 0)
```

## Three ways to draw

The `sample_sets` argument (`scheme` in `draw_sets()`) picks between three
schemes.

**`"balanced"`**, the default. Sets are handed out least-used first, with ties
broken at random. Every choice situation ends up appearing in the sample about
equally often, and the counts never differ by more than one:

```{r balanced}
balanced <- draw_sets(design, respondents = 100, sets_per_resp = 8,
                      scheme = "balanced")
table(table(balanced$Choice.situation))
```

**`"random"`**. Each respondent's sets are drawn independently. Simple, and what
most people picture, but how often each set appears is now itself random:

```{r random}
random <- draw_sets(design, respondents = 100, sets_per_resp = 8,
                    scheme = "random")
range(table(random$Choice.situation))
```

**`"with_replacement"`**. Needed only when `sets_per_resp` exceeds the number of
sets in the design, since then a respondent has to see something twice.

Balance matters because your design was optimised as a whole. If the realised
sample happens to use one choice situation twice as often as another, you are no
longer running the design you chose, and the extra variability shows up in your
estimates for no good reason. `"balanced"` keeps that noise out while leaving the
allocation genuinely random from any individual respondent's point of view.

## Redrawing every run

Here is the part that matters most, and it is easy to get wrong.

Which sets a respondent saw is part of the data generating process, not a fixed
property of the design. So `sim_all()` redraws the allocation for every simulation
run by default. If you drew it once and reused it, you would be measuring how one
particular lucky or unlucky allocation performs, and your standard deviations
across runs would be too small.

```{r resample}
design_dir <- file.path(tempdir(), "ub_designs")
dir.create(design_dir, showWarnings = FALSE)
saveRDS(design, file.path(design_dir, "unblocked.rds"))

ul <- list(
  u1 = list(
    v1 = V.1 ~ bprice * alt1.price + bquality * alt1.quality,
    v2 = V.2 ~ bprice * alt2.price + bquality * alt2.quality
  )
)
bcoeff <- list(bprice = -0.2, bquality = 0.5)

drawn_sets <- function(resample) {
  set.seed(31)
  res <- sim_all(
    nosim = 3, resps = 60, designpath = design_dir, u = ul, bcoeff = bcoeff,
    sets_per_resp = 8, resample = resample,
    estimate = FALSE, mode = "sequential", verbose = 0
  )
  vapply(res$unblocked, function(d) d$Choice_situation[1:5], numeric(5))
}

# default: a different allocation in each run
drawn_sets(resample = TRUE)

# held fixed
drawn_sets(resample = FALSE)
```

Set `resample = FALSE` only when you deliberately want to hold the allocation
constant, for instance to isolate the effect of something else.

## Does it cost you anything?

Random allocation adds a source of variation that blocking removes, so it is fair
to ask what you pay for it. Compare the same design used three ways:

```{r compare}
blocked <- design
blocked$Block <- rep(1:5, each = 8)

blocked_dir <- file.path(tempdir(), "blk_designs")
dir.create(blocked_dir, showWarnings = FALSE)
saveRDS(blocked, file.path(blocked_dir, "blocked.rds"))

compare <- function(path, ...) {
  set.seed(77)
  res <- sim_all(
    nosim = 60, resps = 100, designpath = path, u = ul, bcoeff = bcoeff,
    estimate = TRUE, mode = "sequential", verbose = 0, ...
  )
  nm <- res$arguments$designname
  sa <- res$summaryall
  c(
    price_mean = sa[[paste0(nm, ".mean")]][sa$parname == "bprice"],
    price_sd   = sa[[paste0(nm, ".sd")]][sa$parname == "bprice"],
    qual_mean  = sa[[paste0(nm, ".mean")]][sa$parname == "bquality"],
    qual_sd    = sa[[paste0(nm, ".sd")]][sa$parname == "bquality"]
  )
}

round(cbind(
  blocked  = compare(blocked_dir),
  balanced = compare(design_dir, sets_per_resp = 8),
  random   = compare(design_dir, sets_per_resp = 8, sample_sets = "random")
), 4)
```

All three recover the true coefficients, which is the first thing to check: random
allocation does not bias anything.

The standard deviations across runs are what to compare next. Blocking and balanced
drawing come out essentially the same. Independent random drawing is noticeably
noisier for the price coefficient, because how often each choice situation gets used
now varies from run to run on top of every other source of variation. For the
quality coefficient the three are indistinguishable here. With 60 runs these
standard deviations are themselves uncertain, so treat a small difference as noise
and a large one as real.

The practical summary is that `"balanced"` costs you essentially nothing and buys
you the freedom not to block. `"random"` can cost you precision for no benefit, and
exists mainly because it is what people expect the words "at random" to mean.

## Rules and errors

**Blocked designs are refused.** A `Block` column already says which sets belong
together, so drawing on top of it would undo the blocking:

```{r blocked-error, error = TRUE}
draw_sets(blocked, respondents = 20, sets_per_resp = 4)
```

Drop the column if you want random allocation instead.

**Asking for too many sets is refused**, with the way out named:

```{r too-many, error = TRUE}
draw_sets(design, respondents = 20, sets_per_resp = 50)
```

## Checking the realised allocation

Two things are worth a glance after any run that uses `sets_per_resp`:

```{r checks}
d <- draw_sets(design, respondents = 100, sets_per_resp = 8)

# how evenly the design was used
summary(as.vector(table(d$Choice.situation)))

# nobody saw the same set twice
all(tapply(d$Choice.situation, d$ID, anyDuplicated) == 0)
```

If the first is lopsided you are not running the design you think you are. The
second should always hold unless you asked for `"with_replacement"`.
